Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-09-18 ms.tex
Updating the figures

\begin{abstract}
  We calculate the steady-state hydrodynamic profiles of hot gas in
  the nuclei of a sample of early-type quiescent galaxies from the
  Nuker database.  We assume that mass is supplied to the
  circumnuclear medium (CNM) entirely by stellar winds according to
  the radial stellar profiles for each galaxy, while the gas is heated
  by a combination of shocked stellar winds and supernovae.  Our
  analytic and numerical results are used to estimate the accretion
  rates onto the central supermassive black holes (SMBH) as a function
  of the SMBH mass $M_{\bullet}$ and the heating efficiency, the
  latter of which is related schematically to the age of the stellar
  population, the frequency of supernovae, and the efficacy of
  accretion feedback processes.  For a fixed heating efficiency, we
  find that the SMBH accretion Eddington factor increases with the
  SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto M_{\bullet}^{\alpha}$
  with $\alpha \approx 0.5, 1$ for core and cusp nuclei, respectively.
  % However, considering the potentially lower heating efficiency of low
  % mass SMBHs (due to a younger stellar population and the rareity of
  % supernovae within the stagnation radius) and their resulting higher
  % susceptibility to cooling instability, we find a flatter
  % relationship $\alpha \approx 0$. 
  Our results are used to check previous estimates of the SMBH
  accretion rate assuming Bondi accretion and extrapolating the X-ray
  measured density and temperature to the unresolved Bondi radius.  We
  also discuss the implications of our results for the diversity of
  CNM environments encountered by the jetted outflows from tidal
  disruption events, and for the use of an assumed
  $\dot{M}-M_{\bullet}$ relationship to constrain the occupation
  fraction of SMBHs in low mass galaxies.
\end{abstract}

@@

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by (equation~[\ref{eq:Tanalytic}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 4 \times 10^6\,v_{500}^2 \, \chi   \,\,{\rm K} 
\end{align}
{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Next, by demanding hydrostatic equilibrium at the stagnation point, an expression for the stagnation radius can be derived (equation~[\ref{eq:rstag}]),
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[ (1+\Mstar(\rs)/\Mbh)\left(A-\frac{n}{2}\right)-B\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$ {\bf BDM: note new definition}, $r_{\rm soi} = GM_{\bullet}/\sigma^{2}$ is the radius of the SMBH sphere of influence (soi), and A/B are constants that depend on the Nuker parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where we have assumed $n = 1$ {\bf BDM: true?  should there be a
  $\chi$ in this?}.  Note that this expression is similar to that
obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4} \chi^{-2} \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2} \chi^{-1} \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Note that equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with BH mass at fixed $v_{500}$ but that $\dot{M}$ is extremely sensitive to the assumed heating rate $\propto v_{500}$.

Finally, the density at the stagnation radius $\rho(r_{\rm s})$ can be estimated by using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    3 \times 10^{-24} \Mbheight^{-0.14} v_{500}^{-1} \chi^{-1/2} \eta \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    6 \times 10^{-24}  \Mbheight^{-0.56} v_{500}  \chi^{1/2}  \eta \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it top}), velocity ({\it ADD}), and
    interior X-ray luminosity ({\it bottom}), calculated for three
    characteristic galaxies from our samples: NGC1172 ({\it blue}),
    NGC4478 ({\it green}), and NGC3115 ({\it black}).  Different line
    styles denote different values of the effective wind heating rate,
    $v_{w,0}$=1000 km s$^{-1}$ ({\it solid}) and 200 km s$^{-1}$ ({\it
      dot-dashed}).  {\bf BDM: add velocity profiles...xray luminosity
      can probably go elsewhere}}
\end{figure}

@@

\subsubsection{Millisecond Pulsars} The spin down power of millisecond pulsars (MSPs) could in principle be an important heating source.  If the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each contributes an average spin-down luminosity $L_{\rm sd}$, then the resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm msp}\epsilon$ corresponds to an effective heating rate (equation~ [\ref{eq:vw_eff}])
  \begin{align}
    \vwO=\sqrt{\frac{2 t_h n_{\rm msp} L_{\rm sd}}{\eta}},
  \end{align}
where $\eta < 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$.  If we assume an average spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of Fermi pulsars, then we find that $\vwO\simeq 120\epsilon (\eta/0.1)^{-1/2}$ km s$^{-1}$.  Thus, assuming a high thermalization efficiency $\epsilon \sim 1$, MSP heating can be comparable to main sequence heating.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}, while the binaries giving rise to through accretion MSPs may be disocciated by stellar interactions in the dense nuclear cluster.  
\subsubsection{Type Ia Supernovae}

@@

@@

  \section{Summary}
\label{sec:summary}

>>>>

\begin{abstract}
  We calculate the steady-state hydrodynamic profiles of hot gas in
  the nuclei of a sample of early-type quiescent galaxies.  We assume
  that mass is supplied to the circumnuclear medium (CNM) entirely by
  stellar winds according to the radial stellar profiles for each
  galaxy, while the gas is heated by a combination of shocked stellar
  winds and supernovae.  Our analytic and numerical results are used
  to estimate the accretion rates onto the central supermassive black
  holes (SMBH) as a function of the SMBH mass $M_{\bullet}$ and the
  heating efficiency, the latter of which is related schematically to
  the age of the stellar population, the frequency of supernovae, and
  the efficacy of accretion feedback processes.  For a fixed heating
  efficiency, we find that the SMBH accretion Eddington factor
  increases with the SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto
  M_{\bullet}^{\alpha}$ with $\alpha \approx 0.5, 1$ for cups and core
  nuclei, respectively.
  % However, considering the potentially lower heating efficiency of low
  % mass SMBHs (due to a younger stellar population and the rareity of
  % supernovae within the stagnation radius) and their resulting higher
  % susceptibility to cooling instability, we find a flatter
  % relationship $\alpha \approx 0$. 
  Our results are used to check previous estimates of the SMBH
  accretion rate assuming Bondi accretion and extrapolating the X-ray
  measured density and temperature to the unresolved Bondi radius.  We
  also discuss the implications of our results for the diversity of
  CNM environments encountered by the jetted outflows from tidal
  disruption events, and for the use of an assumed
  $\dot{M}-M_{\bullet}$ relationship to constrain the occupation
  fraction of SMBHs in low mass galaxies.
\end{abstract}

@@

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by (equation~[\ref{eq:Tanalytic}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 6
\times 10^6\,v_{500}^2 \, \mu  \,\,{\rm K} 
\end{align}
%{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Next, by demanding hydrostatic equilibrium at the stagnation point, an expression for the stagnation radius can be derived (equation~[\ref{eq:rstag}]),
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[
    \frac{\Mstar}{\Mbh}\left(A-\frac{n}{2}\right)+(B-\frac{n}{2})\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$, $r_{\rm soi}
=GM_{\bullet}/\sigma^{2}$ is the radius of the SMBH sphere of
influence (soi), and A and B are constants that depend on the Nuker
parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where we have assumed $n = 1$.  Note that this expression is similar to that
obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4}  \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2}  \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Note that equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with BH mass at fixed $v_{500}$ but that $\dot{M}$ is extremely sensitive to the assumed heating rate $\propto v_{500}$.

Finally, the density at the stagnation radius $\rho(r_{\rm s})$ can be estimated by using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    3 \times 10^{-24} \Mbheight^{-0.14} v_{500}^{-1}  \eta \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    6 \times 10^{-24}  \Mbheight^{-0.56} v_{500}    \eta \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it middle}), velocity  ({\it bottom}), and
    interior X-ray luminosity ({\it bottom}), calculated for three
    characteristic galaxies from our samples: NGC1172 ({\it blue}),
    NGC4478 ({\it green}), and NGC3115 ({\it black}).  Different line
    styles denote different values of the effective wind heating rate,
    $v_{w,0}$=1000 km s$^{-1}$ ({\it solid}) and 200 km s$^{-1}$ ({\it
     dot-dashed}).}
\end{figure}

@@

\subsubsection{Millisecond Pulsars} The spin down power of millisecond pulsars (MSPs) could in principle be an important heating source.  If the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each contributes an average spin-down luminosity $L_{\rm sd}$, then the resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm msp}\epsilon$ corresponds to an effective heating rate (equation~ [\ref{eq:vw_eff}])
  \begin{align}
    \vwO=\sqrt{\frac{\epsilon 2 t_h n_{\rm msp} L_{\rm sd}}{\eta}},
  \end{align}
where $\epsilon< 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$.  If we assume an average spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of Fermi pulsars, then we find that $\vwO\simeq 120\epsilon^{1/2} (\eta/0.1)^{-1/2}$ km s$^{-1}$.  Thus, assuming a high thermalization efficiency $\epsilon \sim 1$, MSP heating can be comparable to main sequence heating.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}, while the binaries giving rise to through accretion MSPs may be disocciated by stellar interactions in the dense nuclear cluster.  
\subsubsection{Type Ia Supernovae}

@@

\subsubsection{Comparison with observed X-ray Luminosity-Mass
  Relation}
%Note currently I reverse-engineer the stellar mass from the black
%hole mass inferred from M-sigma; however, it would be better to use
%the tabulated black hole mass from the Mbh-Mbulge relation.
Figure 1 of \citealt{MillerGallo+:2014a} plots the observed
relationship of unresolved x-ray luminosity versus galactic stellar
mass for a sample of quiescent galaxies. We may construct a similar
relationship with our own sample using the inferred accretion rates
from our solutions.  To infer an x-ray luminosity for each solution,
we scale our calculated $\Mdot c^2$ by a constant. We pick $5\times
10^{-7}$, which gives comparable an x-ray luminosities comparable to
those in \citealt{MillerGallo+:2014a}.
%describe the quadratic procedure.

Our results are plotted in Figure~\ref{fig:bh_xray}. For each galaxy
in our sample we select the $\vwO$=200 km/s if $\rs<\rIa$
for this solution,  $\vwO$=500 km/s is $\rs>\rIa$ for this solution,
or exclude it.  $\vwO$ encapsulates heating SNe Ia heating, and it
would only make sense to include this heating if $\rs$>$\rIa$ as
discussed in Section~\ref{sec:heating}. At low $\Mbh$ we have
$\vwO$=200 km/s and at high $\Mbh$ we have $\vwO$=500 km/s.
Note that much of our sample is excluded in this Figure, since for
$\vwO$=200 km/s we would have $\rs>\rIa$ and four $\vwO$=500 km/s we
would have $\rs<\rIa$ (i.e. the relative locations of $\rs$ and $\rIa$
would be inconsistent with the assumed level of heating).

\begin{figure}
\includegraphics[width=\columnwidth]{bh_xray.eps}
\caption{\label{fig:bh_xray} Accretion power plotted as a function of 
  stellar mass for our solutions (points) with the trend of nuclear
  x-ray luminosity with stellar mass and the detection upper limit
  from Figure 1 of \citealt{MillerGallo+:2014a}. In the top panel we
  take the accretion power to be $\epsilon \Mdot c^2$, with
  $\epsilon=5\time 10^{-7}$ chosen to make our accretion powers
  comparable to the observed x-ray luminosities..}
\end{figure}

@@

  \section{Summary and conclusions}
  We have calculated steady-state models for circumnuclear medium of a
  sample of early type galaxies, assuming that gas is supplied
  entirely by stellar wind mass loss and heated by shocked stellar
  winds and SNe Ia. We may use these profiles to calculate an $\Mdot$
  onto each of sample galaxy's SMBH, and get the scaling of $\Mdot$
  with $\Mbh$. We may then compare our results with observed trends of
  nuclear x-ray luminosity. Our main conclusions may be summarized as
  follows.

  \begin{enumerate}
    \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value
      of $\alpha$ depends on the slope of the stellar density profile
      ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively).

  \end{enumerate}
